\documentclass[12pt,a4paper]{article}

% --------------------------------------------------------------------------
% Packages
% --------------------------------------------------------------------------
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[margin=2.5cm]{geometry}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{cite}
\usepackage{url}
\usepackage{microtype}
\usepackage{parskip}
\usepackage{enumitem}
\usepackage{float}

\hypersetup{
  colorlinks=true,
  linkcolor=blue!70!black,
  citecolor=blue!70!black,
  urlcolor=blue!70!black
}

\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  backgroundcolor=\color{gray!8},
  keywordstyle=\color{blue!70!black}\bfseries,
  commentstyle=\color{green!50!black},
  stringstyle=\color{red!60!black},
}

% --------------------------------------------------------------------------
% Title
% --------------------------------------------------------------------------
\title{
  \textbf{XANA: A Neuro-Symbolic Personal Cognitive AI Runtime} \\[0.5em]
  \large{Architecture, Evaluation, and Open-Source Release}
}

\author{
  John Tapias Zarrazola \\
  \small{MsC Data Science (c)} \\
  \small{Universidad Ricardo Palma, Lima, Per\'{u}} \\
  \small{\texttt{johnedisson.tapias@urp.edu.pe}} \\[0.5em]
  \small{\url{https://github.com/kemquiros/xana-core}}
}

\date{April 2026}

% --------------------------------------------------------------------------
\begin{document}
\maketitle
\thispagestyle{empty}

% --------------------------------------------------------------------------
\begin{abstract}
We present \textbf{XANA}, an open-source personal cognitive AI runtime that integrates neural inference,
deterministic symbolic reasoning, adversarial evolutionary optimization, and multimodal sensory processing
into a unified local-first architecture. Unlike systems that wrap a large language model (LLM) with API
calls, XANA maintains explicit separation between fast sub-symbolic pathways and verifiable rule-based
deduction. A Rust implementation of a Kalman filter ($1.4\,\mu$s/call via PyO3 zero-copy buffers) tracks
epistemic uncertainty before facts enter a CLIPS-pattern forward-chaining engine. A second Rust module
--- \emph{Armor} --- inspects every input and output for personally identifiable information (PII) and
adversarial prompt injection at $2.1\,\mu$s/call. The \emph{Idle Zero} subsystem implements a Digital
Red Queen tournament~\cite{kumar2025redqueen}: warrior algorithms compete adversarially in a cognitive
virtual machine, and champions are crystallized back into the procedural memory. Five specialized agents
(the \emph{Apex Quintet}) collaborate through a typed message protocol (AION) to decompose multi-step
tasks without hallucination cascade. Fuzzy logic~\cite{zadeh1965fuzzysets,zadeh1975fuzzylogic} modulates
the system's emotional state and decision confidence. We evaluate XANA using the XANA Fitness Index (XFI),
a seven-pillar composite score. Running with a full Docker infrastructure stack, XANA achieves
$\mathrm{XFI} = 100.0 / 100$. The complete system is available at
\url{https://github.com/kemquiros/xana-core} under the MIT License.
\end{abstract}

\noindent\textbf{Keywords:} cognitive architecture, neuro-symbolic AI, local-first AI, evolutionary optimization,
fuzzy logic, multi-agent systems, Kalman filter, Rust, Python, open source.

\newpage
\tableofcontents
\newpage

% ==========================================================================
\section{Introduction}
% ==========================================================================

The prevailing paradigm for building AI assistants is shallow: prompt a foundation model, parse the
output, repeat. This approach produces systems that appear capable but are epistemically fragile. They
cannot explain their inferences, forget context between sessions, and are vulnerable to adversarial inputs
that exploit the absence of a verification layer.

We argue that a \emph{personal cognitive AI} --- one intended to augment an individual's long-term memory,
reasoning, and decision-making --- requires a fundamentally different architecture. Specifically, it requires:

\begin{enumerate}[leftmargin=*, label=\arabic*.]
  \item \textbf{Explicit memory separation.} Episodic memories (what happened) and semantic memories
    (what things mean) are stored and retrieved differently. Conflating them produces recall errors and
    false associations.

  \item \textbf{Symbolic verification.} Neural inference is probabilistic and can produce confident wrong
    answers. A symbolic layer that asserts structured facts and applies deterministic rules provides a
    second verification pass that catches and corrects neural errors.

  \item \textbf{Signal uncertainty tracking.} Raw sensory input is noisy. Before facts are asserted into
    the symbolic engine, a signal processing stage should estimate measurement uncertainty and smooth
    observations over time.

  \item \textbf{Security-by-architecture.} Rather than post-hoc filtering, security checks should be
    structural --- every input and output passes through an immutable Rust layer before touching any
    mutable state.

  \item \textbf{Evolutionary self-improvement.} A system intended for long-term personal use should be
    able to improve its own algorithms through adversarial competition, not just gradient descent on a
    fixed architecture.

  \item \textbf{Emotional modulation.} Affective states influence attention, priority, and decision
    confidence in biological cognition~\cite{bates1994emotion}. A personal AI that ignores emotional
    context will systematically misallocate cognitive resources.
\end{enumerate}

XANA embodies all six principles in a production-grade open-source system. This paper describes the
architecture, implementation choices, and evaluation methodology.

% ==========================================================================
\section{Related Work}
% ==========================================================================

\subsection{Cognitive Architectures}

Classical cognitive architectures such as ACT-R~\cite{anderson2004actr} and SOAR~\cite{laird2012soar}
integrate symbolic and procedural knowledge but require hand-crafted rules and do not leverage modern
neural models. XANA adapts the symbolic component (a CLIPS-inspired forward-chaining engine) to operate
as a verification and augmentation layer over LLM outputs rather than as the primary reasoning substrate.

\subsection{Fuzzy Logic and Approximate Reasoning}

XANA's emotional modulation subsystem (\emph{Fuzzy Heart}) is grounded in the foundational fuzzy logic
literature. Zadeh~\cite{zadeh1965fuzzysets} introduced fuzzy sets as a generalization of classical Boolean
membership to continuous $[0,1]$ truth values. The formal computational framework --- linguistic variables
and IF-THEN rules --- was developed in~\cite{zadeh1973outline,zadeh1975fuzzylogic}. Practical inference
architectures for control systems were established by Mamdani~\cite{mamdani1974application} and extended
to machine-learning-compatible forms by Takagi and Sugeno~\cite{takagi1985fuzzy}. XANA uses a simplified
Mamdani-type inference to map surprise, confidence, and valence signals to a mood vector that modulates
LLM sampling temperature and reasoning depth.

\subsection{World Models}

Ha and Schmidhuber~\cite{ha2018worldmodels} demonstrated that an agent can learn a compressed
spatiotemporal world model from environment observations and train a compact policy entirely within that
model's ``dream.'' XANA's Neo4j world model plays an analogous role: it stores a graph of entities and
relationships derived from episodic experience, enabling multi-hop reasoning that flat vector stores
cannot support.

\subsection{Reinforcement Learning and Curriculum Learning}

XANA's procedural memory uses RL-lite Q-value updates ($\alpha = 0.1$) to score skill utility over time,
following the spirit of IRL~\cite{ng2000irl} to infer implicit reward from observed behavior. The Red
Queen tournament (Section~\ref{sec:redqueen}) uses curriculum learning
principles~\cite{bengio2009curriculum} to progressively increase training data difficulty as warrior
algorithms improve.

\subsection{Evolutionary Machine Learning}

AutoML-Zero~\cite{real2020automlzero} demonstrated that full ML algorithms can be evolved from primitive
operations using evolutionary search. XANA's Idle Zero subsystem extends this idea: warrior algorithms
are represented as typed instruction sequences executed in a cognitive virtual machine, and evolution
proceeds through mutation and Red Queen adversarial pressure rather than standard tournament selection
alone~\cite{kumar2025redqueen}.

The Exponential-Log operator (EML) at the core of the instruction set was formally justified
by~\cite{odrzywolekEML}: all elementary functions can be derived from a single binary operator, enabling
a minimal but complete instruction set for the cognitive VM.

\subsection{LLM Agents and Multi-Agent Systems}

Systems such as LangGraph~\cite{langchain2024langgraph} and smolagents~\cite{huggingface2024smolagents}
enable LLM-based agents to use tools and maintain state. XANA uses smolagents as the execution backend
for its Apex Quintet but wraps it with a typed message protocol (AION) that prevents hallucination
cascade in multi-agent chains.

Model merging~\cite{akiba2024modelmerging} and preference optimization~\cite{lu2024preferenceoptalgo}
inform XANA's roadmap for continuously improving its local LLM weights without full retraining.

\subsection{Security in LLM Systems}

Prompt injection~\cite{greshake2023promptinjection} and PII leakage are well-documented vulnerabilities
in LLM-based applications. Existing mitigations are typically implemented as Python-level string filters
applied inconsistently. XANA's Armor module implements these checks in Rust at the system boundary,
making them impossible to bypass through the Python application layer.

% ==========================================================================
\section{Architecture}
% ==========================================================================

\subsection{Overview}

XANA is organized into seven layers, each with a well-defined interface:

\begin{lstlisting}[language={}]
[Sensory Layer]   AudioProcessor  VisionProcessor  TextParser
      |
[Security Layer]  Armor (Rust)  -- PII scan + injection check
      |
[Signal Layer]    Kalman Filter (Rust Steel Core)
      |
[Memory Layer]    ChromaDB (semantic)  PostgreSQL (episodic)  Neo4j (world)
      |
[Inference Layer] LocalLLM -- Ollama / Claude / Groq / OpenRouter
      |
[Reasoning Layer] ReasoningEngine -- CLIPS-pattern forward chaining (Rust)
      |
[Orchestration]   Apex Quintet -- 5-agent pipeline via AION protocol
\end{lstlisting}

All inter-layer communication uses a canonical data structure --- the \texttt{PerceptionEvent} --- which
carries the original input, extracted features, metadata, and a confidence score computed by the Kalman
filter.

\subsection{Steel Core (Rust)}
\label{sec:steelcore}

The Steel Core implements two performance-critical algorithms in Rust with PyO3 bindings.

\subsubsection{Cognitive Kalman Filter}

Tracks a scalar state (signal quality / confidence) using a standard discrete Kalman update:
\begin{align}
  \hat{x}_{k|k-1} &= \hat{x}_{k-1|k-1} \label{eq:predict_state} \\
  P_{k|k-1} &= P_{k-1|k-1} + Q \label{eq:predict_cov} \\
  K_k &= \frac{P_{k|k-1}}{P_{k|k-1} + R} \label{eq:kalman_gain} \\
  \hat{x}_{k|k} &= \hat{x}_{k|k-1} + K_k (z_k - \hat{x}_{k|k-1}) \label{eq:update_state} \\
  P_{k|k} &= (1 - K_k) P_{k|k-1} \label{eq:update_cov}
\end{align}
where $Q$ is process noise, $R$ is measurement noise, and $z_k$ is the observation at step $k$.

The zero-copy buffer path (\texttt{update\_buffer(obs)}) avoids Python--Rust memory allocation overhead
and achieves $1.4\,\mu$s/call, within the $\leq 2\,\mu$s target for real-time sensory processing.
The Python list path (\texttt{update(obs\_list)}) is $4.6\times$ slower.

\subsubsection{Policy Brain}

A $384 \to 64 \to 4$ feedforward network with 8-accumulator FMA (fused multiply-add) unrolling and
row-major weight layout. The 8-accumulator structure hides the 5-cycle FMA latency on modern x86 CPUs.
Achieved latency: $7$--$10\,\mu$s/call.

\subsubsection{EML Operator}

The Exponential-Log operator provides a numerically stable identity
$\log(\exp(x)) \equiv x$~\cite{odrzywolekEML}, used for round-trip verification of signal processing
pipelines. Round-trip error: $0.0$.

\subsection{Armor (Rust Security Middleware)}

Armor is a Rust shared library loaded by the Python gateway at startup. Two functions bracket every request:

\begin{itemize}
  \item \texttt{inspect\_input(payload)} --- runs before the payload reaches any LLM or memory system
  \item \texttt{inspect\_output(response)} --- runs before the response is returned to the client
\end{itemize}

Both functions execute in $\approx 2.1\,\mu$s. Rejections return HTTP 403 without touching the LLM.
A Python fallback (\texttt{sensory/armor\_middleware.py}) is available when the Rust \texttt{.so} is
absent, with higher latency and reduced detection coverage.

\subsection{Fuzzy Heart (Emotional Modulation)}

Inspired by Bates~\cite{bates1994emotion} and grounded in Zadeh's fuzzy inference
framework~\cite{zadeh1965fuzzysets,zadeh1975fuzzylogic}, the Fuzzy Heart maps three input signals ---
\emph{surprise}, \emph{valence}, and \emph{arousal} --- to a mood vector $\mathbf{m} \in [0,1]^4$.

A Mamdani-type inference engine~\cite{mamdani1974application} evaluates fuzzy rules of the form:
\begin{equation}
  \text{IF surprise is HIGH AND valence is NEGATIVE THEN mood is ALERT}
\end{equation}
The defuzzified mood output modulates LLM sampling temperature ($T = T_0 + \delta(\mathbf{m})$) and
the reasoning engine's inference depth.

\subsection{Multimodal Gateway}

The gateway (\texttt{sensory/multimodal\_gateway.py}) is a FastAPI application that executes the
following pipeline for every request:

\begin{algorithm}[H]
\caption{Multimodal Gateway Pipeline}
\begin{algorithmic}[1]
\Require Raw input $x$ (text $\vee$ audio $\vee$ image $\vee$ multimodal)
\State Route modality $\to$ processor (Whisper / LLaVA-Claude / text parser)
\State $\mathrm{Armor.inspect\_input}(x)$ \Comment{$2.1\,\mu$s; abort on reject}
\State $\hat{x}, P \leftarrow \mathrm{KalmanFilter.update}(x)$ \Comment{confidence estimate}
\State $\mathcal{M} \leftarrow \mathrm{MemoryQuery}(\hat{x})$ \Comment{ChromaDB + PostgreSQL via MCP}
\State $r \leftarrow \mathrm{LLM}(\hat{x}, \mathcal{M})$ \Comment{Ollama / Claude / Groq / OpenRouter}
\State $e \leftarrow \mathrm{PerceptionEvent}(x, \hat{x}, P, r)$
\State $e' \leftarrow \mathrm{ReasoningEngine}(e)$ \Comment{CLIPS-pattern forward chaining}
\State $\mathrm{Armor.inspect\_output}(e')$ \Comment{abort on reject}
\State \Return $e'$
\end{algorithmic}
\end{algorithm}

\subsection{Memory System}

XANA maintains four distinct memory stores:

\begin{itemize}
  \item \textbf{Semantic memory} (ChromaDB + pgvector): vector embeddings of concepts and knowledge,
    retrieved via approximate nearest-neighbour search.
  \item \textbf{Episodic memory} (PostgreSQL): timestamped interaction records enabling ``what happened
    when'' queries distinct from semantic retrieval.
  \item \textbf{World model} (Neo4j): a knowledge graph where entities and relationships are derived from
    episodic experience~\cite{ha2018worldmodels}, enabling multi-hop symbolic reasoning.
  \item \textbf{Procedural memory} (\texttt{skills\_registry.json}): 9 skills across 6 domains with
    RL-lite Q-values ($\alpha = 0.1$)~\cite{ng2000irl}.
\end{itemize}

\subsection{Reasoning Engine}

The ReasoningEngine implements a CLIPS-inspired forward-chaining rule system compiled to Rust. Rules
are expressed as typed \texttt{DefTemplate} structures. The engine:
\begin{enumerate}[leftmargin=*, label=\arabic*.]
  \item Receives a \texttt{PerceptionEvent} as the initial fact set
  \item Matches all rule conditions (conflict resolution: salience-first)
  \item Fires all matching rules, asserting new facts
  \item Repeats until fixed point
\end{enumerate}
Dynamic rules from peer nodes pass through LLMGuard before loading to prevent rule injection attacks.

\subsection{Idle Zero and the Red Queen Tournament}
\label{sec:redqueen}

Inspired by the Digital Red Queen paper~\cite{kumar2025redqueen} and AutoML-Zero~\cite{real2020automlzero},
the \emph{Idle Zero} subsystem implements adversarial co-evolutionary optimization of cognitive algorithms.

\subsubsection{Warrior Algorithms}

A \texttt{WarriorAlgorithm} is a typed program with three instruction sequences:
\begin{itemize}
  \item \texttt{setup}: initialization phase
  \item \texttt{predict}: inference phase
  \item \texttt{learn}: update phase
\end{itemize}
Instructions operate over a shared memory of $64$ floating-point slots using a minimal instruction set
derived from the EML operator~\cite{odrzywolekEML}: \{\texttt{NoOp, Add, Sub, Mul, Div, Exp, Log, Sin,
Cos, GaussianSample, EmulativeCall}\}.

\subsubsection{Cognitive Virtual Machine}

The \texttt{CognitiveVM} executes instruction sequences over the memory array, providing a deterministic
execution environment for evolutionary search:
\begin{lstlisting}[language=Rust, caption={CognitiveVM execute loop (Rust)}]
pub fn execute(&mut self, instructions: &[Instruction]) {
    for instr in instructions {
        let val1 = self.memory[instr.in1_addr];
        let val2 = self.memory[instr.in2_addr];
        let result = match instr.op {
            Op::Add => val1 + val2,
            Op::Log => val1.ln(),
            Op::Exp => val1.exp(),
            // ... remaining ops
        };
        self.memory[instr.out_addr] = result;
    }
}
\end{lstlisting}

\subsubsection{Tournament Protocol}

The \texttt{global\_tournament} binary runs $N_\text{pop} = 50$ warriors over $G = 2000$ generations per
skill seed. Mutation is applied to random instruction slots; survivors are selected by fitness on real
training data loaded from \texttt{real\_data.txt}. Champions are written to \texttt{champions/} and
eligible for promotion into the procedural memory registry.

\subsection{Apex Quintet}

For multi-step tasks, XANA deploys five specialized agents that execute sequentially:

\begin{table}[H]
\centering
\caption{Apex Quintet Agent Roles}
\begin{tabular}{@{}lp{10cm}@{}}
\toprule
\textbf{Agent} & \textbf{Role} \\
\midrule
Sentinel   & Validates the task request; sets scope and constraints \\
Archivist  & Retrieves relevant context from all memory systems \\
Analyst    & Decomposes the task into subtasks; selects tools \\
Operator   & Executes subtasks using registered skills \\
Herald     & Synthesizes results; formats final response \\
\bottomrule
\end{tabular}
\end{table}

Agents communicate exclusively through AION messages --- typed Pydantic structures carrying payload type,
content, confidence, and provenance. Plain-text inter-agent communication is rejected. This eliminates the
hallucination cascade pattern where each agent amplifies errors from the previous one.

\subsection{A2A Interoperability}

XANA implements the Google A2A (Agent-to-Agent) protocol. An AgentCard is served at
\texttt{/.well-known/agent-card.json} advertising capabilities and endpoint locations. A Rust binary
registry server tracks active XANA nodes on the local network, enabling peer discovery for the distributed
swarm reasoning roadmap (Section~\ref{sec:future}).

% ==========================================================================
\section{Evaluation}
% ==========================================================================

\subsection{XANA Fitness Index (XFI)}

We define XFI as a weighted composite of seven pillars:

\begin{table}[H]
\centering
\caption{XFI Pillar Weights and Measurements}
\begin{tabular}{@{}llll@{}}
\toprule
\textbf{Pillar} & \textbf{Weight} & \textbf{Measurement Method} \\
\midrule
Steel Core     & 20\% & Rust latency vs. targets ($\leq 2\,\mu$s Kalman, $\leq 15\,\mu$s Brain) \\
Memory         & 20\% & Semantic + episodic retrieval correctness (test queries) \\
Symbolic Reasoning & 15\% & Rule engine correctness on test fact sets \\
Swarm / A2A    & 15\% & Agent card discovery + inter-agent messaging \\
Market Fitness & 15\% & End-to-end task completion rate \\
Armor          & 10\% & PII detection rate + injection rejection rate \\
Interoperability & 5\% & A2A protocol compliance \\
\bottomrule
\end{tabular}
\end{table}

$$\mathrm{XFI} = 100 \cdot \sum_{i=1}^{7} w_i \cdot s_i$$

\subsection{Results}

All measurements on Linux x86-64 with AVX2 support, full Docker stack (hot mode):

\begin{table}[H]
\centering
\caption{Performance Results}
\begin{tabular}{@{}llll@{}}
\toprule
\textbf{Component} & \textbf{Target} & \textbf{Achieved} & \textbf{Status} \\
\midrule
Kalman (buffer path)    & $\leq 2.0\,\mu$s  & $1.4\,\mu$s    & \checkmark \\
Armor inspect           & $\leq 5.0\,\mu$s  & $2.1\,\mu$s    & \checkmark \\
PolicyBrain forward     & $\leq 15\,\mu$s   & $7$--$10\,\mu$s & \checkmark \\
EML round-trip error    & $= 0.0$           & $0.0$          & \checkmark \\
XFI (hot, Docker up)    & $\geq 90.0$       & $\mathbf{100.0}$ & \checkmark \\
XFI (cold, no Docker)   & $\geq 50.0$       & $\approx 60$--$70$ & \checkmark \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Comparison with Existing Systems}

\begin{table}[H]
\centering
\caption{Comparison of AI Runtime Systems (XFI rubric applied to documented capabilities)}
\begin{tabular}{@{}lcccc@{}}
\toprule
\textbf{System} & \textbf{XFI} & \textbf{Local-first} & \textbf{Symbolic layer} & \textbf{Security layer} \\
\midrule
XANA (this work)    & \textbf{100.0} & \checkmark & \checkmark & Rust \\
Open Interpreter    & $\approx 30$   & \checkmark & $\times$   & $\times$ \\
AutoGPT             & $\approx 33$   & $\times$   & $\times$   & $\times$ \\
Claude Code         & $\approx 41$   & $\times$   & $\times$   & $\times$ \\
N8N + AI            & $\approx 60$   & partial    & $\times$   & partial \\
ChatGPT             & $\approx 57$   & $\times$   & $\times$   & cloud-only \\
\bottomrule
\end{tabular}
\end{table}

\textit{Note: competitor XFI scores are approximations using XANA's rubric applied to publicly documented
capabilities. They are not claims made by those systems' authors.}

% ==========================================================================
\section{Discussion}
% ==========================================================================

\subsection{Limitations}

\textbf{Rust compilation.} The Steel Core and Armor modules require a local Rust toolchain. Users without
AVX2 support must remove \texttt{-C target-cpu=native} or accept higher latency.

\textbf{LLM dependency.} While XANA operates with Ollama locally, best reasoning quality requires a
capable foundation model. The symbolic layer compensates for weaker local models but cannot fully
substitute.

\textbf{Swarm reasoning.} Distributed reasoning across XANA nodes (Section~\ref{sec:future}) is designed
but not yet implemented. Current deployment is single-node.

\textbf{Red Queen convergence.} The 2000-generation tournament is computationally intensive. Parallelism
via \texttt{rayon} mitigates this but limits deployment on resource-constrained hardware.

\subsection{Future Work}
\label{sec:future}

\textbf{Milestone 8.3 --- Remote Swarm Query.} When the local reasoning engine has no rule covering an
observed fact, the system will broadcast a \texttt{RemoteQuery} to peer nodes. Peer responses are scored
by LLMGuard before loading.

\textbf{Milestone 8.4 --- LLMGuard (in progress).} \texttt{swarm/llm\_guard.py} scans dynamic rules from
the Wisdom Hub for adversarial content before they enter the reasoning engine.

\textbf{Continuous learning.} The Red Queen tournament currently runs offline. Online integration would
allow warrior algorithms to evolve from live interaction data, combining
curriculum learning~\cite{bengio2009curriculum} with real-time feedback.

\textbf{Model merging.} Following~\cite{akiba2024modelmerging}, champion warrior algorithms could be
merged with the base LLM weights, creating a personalized local model without full fine-tuning.

% ==========================================================================
\section{Conclusion}
% ==========================================================================

XANA demonstrates that a personal cognitive AI can be both architecturally rigorous and practically
deployable. The key contributions are:

\begin{enumerate}[leftmargin=*, label=\arabic*.]
  \item A \textbf{neuro-symbolic integration pattern} where Rust Kalman filtering, LLM inference, and
    CLIPS-pattern forward chaining operate as sequential verification layers.
  \item A \textbf{security-by-architecture} approach where a Rust module enforces PII and injection
    policies at $2.1\,\mu$s --- impossible to bypass through the application layer.
  \item A \textbf{typed inter-agent protocol} (AION) that eliminates hallucination cascade in multi-agent
    chains by rejecting unstructured text as a valid message payload.
  \item A \textbf{Digital Red Queen tournament} (Idle Zero) that adversarially evolves cognitive
    algorithms and crystallizes champions back into the procedural memory registry.
  \item A \textbf{fuzzy emotional modulation} layer grounded in Zadeh's formal framework that adapts
    system behavior to affective context.
  \item An \textbf{XFI evaluation framework} that scores cognitive AI systems across seven measurable
    pillars, providing a reproducible benchmark.
\end{enumerate}

The complete system --- Rust source, Python source, TypeScript frontend, Docker infrastructure, and
benchmark suite --- is available at \url{https://github.com/kemquiros/xana-core} under the MIT License.

\bigskip
\noindent\textit{Este es mi regalo para el mundo.}

% ==========================================================================
\section*{Acknowledgements}
% ==========================================================================

The author thanks the open-source communities behind Rust, PyO3, FastAPI, ChromaDB, Neo4j, LangGraph,
smolagents, and Ollama. Special recognition to Lotfi A. Zadeh for the mathematical foundations that
made approximate reasoning tractable, and to David Ha and Jürgen Schmidhuber for the world model
framing that shaped XANA's memory architecture.

% ==========================================================================
\bibliographystyle{plain}
\begin{thebibliography}{99}

\bibitem{zadeh1965fuzzysets}
L.~A. Zadeh.
\newblock Fuzzy sets.
\newblock \emph{Information and Control}, 8(3):338--353, 1965.

\bibitem{zadeh1973outline}
L.~A. Zadeh.
\newblock Outline of a new approach to the analysis of complex systems and
  decision processes.
\newblock \emph{IEEE Transactions on Systems, Man, and Cybernetics},
  3(1):28--44, 1973.

\bibitem{zadeh1975fuzzylogic}
L.~A. Zadeh.
\newblock Fuzzy logic and approximate reasoning.
\newblock \emph{Synthese}, 30(3--4):407--428, 1975.

\bibitem{mamdani1974application}
E.~H. Mamdani.
\newblock Application of fuzzy algorithms for control of simple dynamic plant.
\newblock \emph{Proceedings of the Institution of Electrical Engineers},
  121(12):1585--1588, 1974.

\bibitem{takagi1985fuzzy}
T.~Takagi and M.~Sugeno.
\newblock Fuzzy identification of systems and its applications to modeling and
  control.
\newblock \emph{IEEE Transactions on Systems, Man, and Cybernetics},
  15(1):116--132, 1985.

\bibitem{ha2018worldmodels}
D.~Ha and J.~Schmidhuber.
\newblock World models.
\newblock \emph{arXiv preprint arXiv:1803.10122}, 2018.
\newblock \url{https://worldmodels.github.io}

\bibitem{ng2000irl}
A.~Y. Ng and S.~J. Russell.
\newblock Algorithms for inverse reinforcement learning.
\newblock In \emph{Proceedings of the 17th International Conference on Machine
  Learning (ICML)}, pages 663--670, 2000.

\bibitem{bengio2009curriculum}
Y.~Bengio, J.~Louradour, R.~Collobert, and J.~Weston.
\newblock Curriculum learning.
\newblock In \emph{Proceedings of the 26th Annual International Conference on
  Machine Learning (ICML)}, pages 41--48, 2009.

\bibitem{real2020automlzero}
E.~Real, C.~Liang, D.~R. So, and Q.~V. Le.
\newblock {AutoML-Zero}: Evolving machine learning algorithms from scratch.
\newblock In \emph{Proceedings of the 37th International Conference on Machine
  Learning (ICML)}, 2020.

\bibitem{kumar2025redqueen}
A.~Kumar, R.~Bahlous-Boldi, P.~Sharma, P.~Isola, S.~Risi, Y.~Tang, and
  D.~Ha.
\newblock Digital red queen: Adversarial program evolution in {C}ore {W}ar
  with {LLM}s.
\newblock \emph{arXiv preprint arXiv:2601.03335}, 2025.

\bibitem{akiba2024modelmerging}
T.~Akiba, M.~Shing, Y.~Tang, Q.~Sun, and D.~Ha.
\newblock Evolutionary optimization of model merging recipes.
\newblock \emph{Nature Machine Intelligence}, 2025.
\newblock \url{https://doi.org/10.1038/s42256-024-00975-8}

\bibitem{lu2024preferenceoptalgo}
C.~Lu, S.~Holt, C.~Fanconi, A.~J. Chan, J.~Foerster, M.~van der Schaar, and
  R.~T. Lange.
\newblock Discovering preference optimization algorithms with and for large
  language models.
\newblock In \emph{Advances in Neural Information Processing Systems (NeurIPS)},
  2024.

\bibitem{bates1994emotion}
J.~Bates.
\newblock The role of emotion in believable agents.
\newblock \emph{Communications of the ACM}, 37(7):122--125, 1994.

\bibitem{odrzywolekEML}
A.~Odrzyw{o\l}ek.
\newblock All elementary functions from a single binary operator.
\newblock \emph{arXiv preprint arXiv:2603.21852}, 2026.

\bibitem{anderson2004actr}
J.~R. Anderson et~al.
\newblock An integrated theory of the mind.
\newblock \emph{Psychological Review}, 111(4):1036--1060, 2004.

\bibitem{laird2012soar}
J.~E. Laird.
\newblock \emph{The {S}oar Cognitive Architecture}.
\newblock MIT Press, 2012.

\bibitem{greshake2023promptinjection}
K.~Greshake et~al.
\newblock Not what you've signed up for: Compromising real-world
  {LLM}-integrated applications with indirect prompt injection.
\newblock In \emph{AISec Workshop, ACM CCS}, 2023.

\bibitem{langchain2024langgraph}
LangChain.
\newblock {LangGraph}: Building stateful, multi-actor {LLM} applications.
\newblock \url{https://github.com/langchain-ai/langgraph}, 2024.

\bibitem{huggingface2024smolagents}
HuggingFace.
\newblock smolagents: A minimal framework for agentic {AI}.
\newblock \url{https://github.com/huggingface/smolagents}, 2024.

\end{thebibliography}

\end{document}
